---
title: "Motor Trend MPG Data Analysis"
author: "Jody9678"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
  *Created with knitr*

#### Executive Summary
This report analyzes the relationship between transmission type (manual or automatic)
and miles per gallon (MPG) as a means to answer two questions:
  Which transmission type is better for MPG?
  What is the MPG difference between automatic and manual transmissions?

The `mtcars` dataset was used for this analysis. A simple t-test between
automatic and manual transmission vehicles indicates that vehicles with manual
transmissions are far more efficient, achieving 7.24 more miles per gallon than
their automatic counterparts.
A deeper dive through the application of multiple linear regressions provided
a more meaningful view of the data and a more concise conclusion; in that the
transmission type contributed less significantly to actual fuel efficiency
(manual transmissions over automatic by a mere 1.81 MPG) with factors of weight,
horsepower, and number of cylinders contributing a great deal more to the overall
fuel efficiency of vehicles.
Manual transmission is the overall winner on efficiency, but size and power
change the picture significantly.

#### Load Data
Load the dataset and convert categorical variables to factors.
```{r results='hide', message=FALSE}
library(ggplot2)
data(mtcars)
head(mtcars, n=3)
dim(mtcars)
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$vs <- as.factor(mtcars$vs)
mtcars$am <- factor(mtcars$am)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
attach(mtcars)
```

#### Exploratory Analysis

**See Appendix Figure I** 
'Exploratory Box graph comparing Automatic and Manual transmission MPG.'

The graph leads us to believe that there is a significant improvement in MPG
for vehicles with a manual transmission vs automatic.

##### Statistical Inference
T-Test transmission type and MPG
```{r}
testResults <- t.test(mpg ~ am)
testResults$p.value
```
The T-Test rejects the null hypothesis that the difference between transmission
types is 0.
```{r}
testResults$estimate
```
The difference estimate between the 2 transmissions is 7.24494 MPG in favor of
manual transmissions.

##### Regression Analysis
Fit the full model of the data
```{r results='hide'}
fullModelFit <- lm(mpg ~ ., data = mtcars)
summary(fullModelFit)  # results hidden
summary(fullModelFit)$coeff  # results hidden
```
Since none of the coefficients have a p-value less than 0.05 we cannot conclude
which variables are more statistically significant.

Backward selection to determine which variables are most statistically
significant
```{r results='hide'}
stepFit <- step(fullModelFit)
summary(stepFit) # results hidden
summary(stepFit)$coeff # results hidden
```

The new model has 4 variables (cylinders, horsepower, weight, transmission). The
R-squared value of 0.8659 confirms that this model explains almost 87% of the
variance in MPG. The p-values also are statistically significantly because they
have a p-value less than 0.05. The coefficients conclude that increasing the
number of cylinders from 4 to 6 with decrease the MPG by 3.03, with an additional
drop 0f 2.16 MPG with 8 cylinder vehicles.
Efficiency drops by 3.21 MPG for every 100 horsepower added.
Every 1000lb increase costs 2.5 MPG.
Taking all of this into account, manual transmissions offer a marginal 1.81 MPG 
improvement over automatic transmissions.

#### Residuals & Diagnostics

**See Appendix Figure II**
'Residual Plots including all available variables.'

  The plots conclude:

  1. The randomness of the Residuals vs. Fitted plot supports the assumption of
independence
2. The points of the Normal Q-Q plot following closely to the line conclude that
the distribution of residuals is normal
3. The Scale-Location plot random distribution confirms the constant variance
assumption
4. Since all points are within the 0.05 lines, the Residuals vs. Leverage
concludes that there are no outliers
```{r}
sum((abs(dfbetas(stepFit)))>1)
```

#### Conclusion
There is a difference in MPG based on transmission type, with manual transmissions
offering a slight improvement in fuel efficiency. However, it seems that weight,
horsepower, & number of cylinders are more statistically significant when
determining MPG.

### Appendix Figures

#### Figure I
```{r echo=FALSE}
boxplot(mpg ~ am,
        xlab="Transmission Type (0 = Automatic, 1 = Manual)",
        ylab="MPG",
        main="MPG by Transmission Type")
```
'Exploratory Box graph comparing Automatic and Manual transmission MPG.'


#### Figure II
```{r echo=FALSE}
par(mfrow = c(2, 2))
plot(stepFit)
```
'Residual Plots including all available variables.'
